LCS Tool: A Computational Platform for Lagrangian Coherent Structures

### 2014-12-12 main.tex

\documentclass[5p]{elsarticle}

\usepackage{amsmath}
\usepackage{commath}
\usepackage{gensymb}
\usepackage{import}
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage[numbered,framed]{mcode}
\usepackage{microtype}
\usepackage{pgfplots}
\usepackage[hidelinks]{hyperref}
\usepackage{cleveref}

@@

\lstloadlanguages{Matlab}
\lstset{
  basicstyle=\small,
  breaklines,
  breakatwhitespace,
  frame=single,
  float,
  columns=fullflexible,
  numbers=left,
  stepnumber=2,
  backgroundcolor=\color{grey},
  escapeinside={\%@}{\^^M},
  mathescape
}

% Workaround for bug in elsarticle-num-names.bst
% Reference: http://www.tug.org/pipermail/tex-live/2012-August/032152.html
\makeatletter
\providecommand{\doi}[1]{%
  \begingroup
    \let\bibinfo\@secondoftwo
    \urlstyle{rm}%
    \href{http://dx.doi.org/#1}{%
      doi:\discretionary{}{}{}%
      \nolinkurl{#1}%
    }%
  \endgroup
}
\makeatother

@@

\begin{sloppypar}
Lagrangian Coherent Structures (LCSs) are evolving organizing centres of trajectory patterns in non-autonomous dynamical systems\citep{haller00:_lagran,peacock13:_lagran,haller15:_langr_coher_struc}. Applications of LCSs include oceanic and atmospheric flows\citep{beron-vera13:_objec_agulh,koh02:_hyper}, biological transport problems\citep{wilson09:_lagran_reynol,tallapragada11:_lagran,huhn12:_south_indian_ocean_count_madag}, aeronautics\citep{tang10:_accur_lagran_hong_kong_inter_airpor}, celestial mechanics\citep{gawlik09:_lagran}, crowd dynamics\citep{ali07:_lagran_partic_dynam_approac_crowd}, and aperiodically forced mechanical oscillators\citep{hadjighasem13:_detec_kam}.
\end{sloppypar}

\citet{haller01:_distin} proposed that ridges of the finite-time Lyapunov exponent (FTLE) are heuristic indicators of hyperbolic (i.e., repelling and attracting type) LCSs.
A number of examples support this principle\citep{peacock13:_lagran}.
Equating FTLE ridges with LCSs, however, would create theoretical inconsistencies, as well as false positives and negatives in hyperbolic LCS detection\citep{haller11:_lagran_coher_struc,norgard12:_secon_lagran_coher_struc}.
In addition, the role of the FTLE field in the accurate detection of elliptic (vortex-type) and parabolic (jet-core type) LCSs has remained an open question (but see \citet{beron-vera10:_invar_lagran}).

More recent work has focused on an exact mathematical formulation of the properties defining LCSs\citep{haller11:_lagran_coher_struc,farazmand12:_comput_lagran,haller12:_geodes_theor_trans_barrier_two_dimen_flows,haller13:_coher_lagran,haller14:_adden_coher_lagran,farazmand13:_attrac_lagran,blazevski14:_hyper_ellip_trans_barrier_three}.
In two-dimensional flows, hyperbolic and parabolic LCSs turn out to be stationary curves of the averaged material shear\citep{farazmand14:_shearless}, whereas elliptic LCSs are stationary curves of the averaged strain\citep{haller13:_coher_lagran,haller14:_adden_coher_lagran}.
These variational formulations lead to explicit solutions for LCSs as null geodesics of appropriate Lorentzian metrics.

\begin{sloppypar}
We present an algorithmic introduction to geodesic LCS detection.
We then review the implementation of this approach in a computational engine called LCS Tool\footnote{LCS Tool is available at: \url{http://www.runmycode.org/companion/view/908}}.
This engine is a library of MATLAB functions that extract LCSs from two-dimensional unsteady flows. The examples we present form demonstration scripts distributed with LCS Tool.
\end{sloppypar}

Other software for LCS detection exists.
ManGen\citep{lekien03:_time} calculates the FTLE and advects material curves in two-dimensional velocity fields.
It includes a graphical user interface and uses the Message Passing Interface standard for parallel calculations.
Newman\citep{toit10:_trans} calculates the FTLE using dimension independent code.
It assists ridge extraction of FTLE fields and supports analytic and dataset velocity definitions.
FlowTK\citep{ameli14:_devel_effic_flexib_pipel_lagran} calculates the FTLE in two and three-dimensions on Cartesian and unstructured grids.
The NVIDIA CUDA parallel computing platform is used for fast computations and Kitware's ParaView data analysis and visualization application is used for a user interface.

@@

Next we consider positions of material lines at time $t_0$ that prevail as most repelling or attracting material lines (or \emph{hyperbolic LCSs}) over a time interval $[t_0,t] \subset [t_-,t_+]$. \citet{farazmand14:_shearless} argue that hyperbolic LCSs are stationary curves of the averaged shear functional
\begin{gather*}
Q(\gamma) = \frac{1}{\sigma} \int_0^\sigma \frac{\langle r'(s),D_{t_0}^t(r(s)) r'(s)\rangle}{\sqrt{\langle r'(s),C_{t_0}^t(r(s)) r'(s)\rangle\langle r'(s),r'(s)\rangle}}\dif s,\\
D_{t_0}^t = \frac12 [C_{t_0}^t \Omega - \Omega C_{t_0}^t],
\end{gather*}
obtained by averaging the Lagrangian shear arising over $[t_0,t]$ along material lines parametrized as $r(s)$ with $s \in [0,\sigma]$.
More precisely, hyperbolic LCSs are stationary curves of $Q(\gamma)$ with respect to fixed-endpoint perturbations.
We note that parabolic LCSs (Lagrangian jet cores) are also stationary curves of $Q(\gamma)$, but under variable endpoint perturbations (cf. \citet{farazmand14:_shearless}).

@@

Shrink lines and stretch lines coincide with the null geodesics of the Lorentzian metric $h(u,v) = \left\langle u,D_{t_0}^t(r) v \right\rangle$.
For this reason, we refer to the computation of hyperbolic LCSs as strongest normally-repelling or normally-attracting curves of \cref{eq:etafields} as \emph{geodesic detection of hyperbolic LCSs}.

@@

\begin{table}
\begin{enumerate}
\item Define a velocity field, \lstinline!dx = derivative(t,x,p)!
\item Compute Cauchy-Green strain tensor invariants, \lstinline![v,d] = eig_cgStrain(derivative)!
\item Define a range of $\lambda$ values, \lstinline!lambda!
\item Define Poincare sections, \lstinline!ps!
\item Detect elliptic LCSs, \lstinline!closedLambdaLine = poincare_closed_orbit_range(v,d,lambda,ps)! and \lstinline!ellipticLcs = elliptic_lcs(closedLambdaLine)!
\item Define a local maximization/minimization distance for hyperbolic LCSs, \lstinline!lmd!
\item Detect hyperbolic LCSs, \lstinline!hyperbolicLine = seed_curves_from_lambda_max(v,d,ellipticLcs,lmd)! and \lstinline!hyperbolicLcs = remove_strain_in_elliptic( hyperbolicLine,ellipticLcs)!
\end{enumerate}
\caption{Overview of sequence of computations to detect LCSs with LCS Tool functions.}
\label{t:LCS Tool overview}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Resolution & Runtime (s) & Memory (MB)\tabularnewline
\hline \hline
$500 \times 251$ & 29 & 55\tabularnewline
\hline
$750 \times 376$ & 63 & 123\tabularnewline
\hline
$1000 \times 501$ & 112 & 223\tabularnewline
\hline
\end{tabular}
\end{center}
\caption{
Runtime and memory use associated with vector integration at different resolutions.
Results are from the function \lstinline!eig_cgStrain! applied to the double gyre presented in \cref{sec:double gyre}.
Results obtained with an Intel Core i3-4130 3.4 GHz processor.
}
\label{t:vector integration memory}
\end{table}

The Cauchy-Green strain tensor can have point singularities, i.e., points where $C_{t_0}^t(x_0)$ has repeated eigenvalues.
At these points the eigenvectors $\xi_1(x_0)$ and $\xi_2(x_0)$ are no longer well-defined.
This generically arises at a finite set of isolated points within the computational domain\citep{delmarcelle94}, and hence lie off the computational grid with probability one.

@@

\begin{table*}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline
\multicolumn{2}{|p{.95\textwidth}|}
{\lstinline![shearline.etaPos,shearline.etaNeg] = lambda_line(cgEigenvector,cgEigenvalue,lambda)!}\tabularnewline
\hline \hline
\lstinline!cgEigenvector! & array of Cauchy-Green strain eigenvectors\tabularnewline
\hline
\lstinline!cgEigenvalue! & array of Cauchy-Green strain eigenvalues\tabularnewline
\hline
\lstinline!lambda! & scalar lambda value in \cref{eq:eta}\tabularnewline
\hline \hline
\multicolumn{2}{|p{.95\textwidth}|}{\lstinline![closedOrbits,orbits] = poincare_closed_orbit_multi(domain,resolution,shearline,PSList)!}\tabularnewline
\hline \hline
\lstinline!domain! & array to define flow domain\tabularnewline
\hline
\lstinline!resolution! & $1 \times 2$ array to define main grid resolution for Cauchy-Green strain tensor\tabularnewline
\hline
\lstinline!shearline! & structure of arrays of $\eta_+$ and $\eta_-$ values on main grid\tabularnewline
\hline
\lstinline!PSList! & user-defined structure for Poincare section end-points, number of $\lambda$-lines launched from Poincare section, and maximum closed $\lambda$-line length\tabularnewline
\hline
\lstinline!nBisection! & optional number of bisection steps to refine zero crossings of Poincare map. Default: 5.\tabularnewline
\hline
\lstinline!dThresh! & optional threshold to discard discontinuous zero crossings of Poincare map. Default: $10^{-2}$.\tabularnewline
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify $\lambda$-line integration parameters\tabularnewline
\hline
\lstinline!periodicBc! & optional $1 \times 2$ logical array to specify periodic boundary conditions. Default: \lstinline![false,false]!.\tabularnewline
\hline
\end{tabular}
\caption{
Syntax for LCS Tool elliptic LCS functions.
Both functions are called by the function \lstinline!poincare_closed_orbit_range!.
}
\label{t:Elliptic LCS functions}
\end{table*}

The first step is to define the position of Poincare sections in regions where closed $\lambda$-lines are expected based on a visual analysis of the orbit structure of the $\eta_\pm^\lambda$ vector field.
The Poincare section is to be oriented such that the first endpoint is close to the centre of the expected Lagrangian vortex, and the second endpoint is outside this vortex.
Additionally, the number of lambda-lines launched from the Poincare section, \lstinline!poincareSection.numPoints!, must be defined.
A reasonable default value is 100.

@@

\begin{table}
\begin{center}
\begin{tabular}{|p{.3\textwidth}|c|}
\hline
Demo & Runtime (min)\tabularnewline
\hline \hline
double gyre & 50\tabularnewline
\hline
Bickley jet & 975\tabularnewline
\hline
ocean dataset & 127\tabularnewline
\hline
\end{tabular}
\end{center}
\caption{
Runtime of some demo scripts.
Results obtained with an Intel Core i3-4130 3.4 GHz processor with 8 GB of memory.
}
\label{t:demo runtimes}
\end{table}

@@

The double gyre is a model for a time-dependent two gyre system observed in geophysical flows \citep{shadden05:_defin_lagran_lyapun}. The model consists of two counter rotating sinusoidal vortices with a harmonically oscillating line in-between. Lagrangian particle motions satisfy the non-autonomous dynamical system
\begin{equation}
\begin{split}
\od{x}{t} = -\pi A \sin[\pi f(x,t)] \cos(\pi y),\\
\od{y}{t} = \pi A \cos[\pi f(x,t)] \sin(\pi y) \pd{f(x,t)}{x},\\
f(x,t) = \epsilon \sin(\omega t) x^2 + [1 - 2 \epsilon \sin(\omega t)] x.
\end{split}
\label{eq:double gyre derivative equations}
\end{equation}
The MATLAB function describing this velocity field is given in \cref{l:double gyre derivative}, specifying the right hand side of the particle ordinary differential equation in a way that supports vectorized integration.

@@

In what follows, the parameter values are: $A = 0.1$, $\epsilon = 0.1$, $\omega = \pi/5$.
The flow timespan is $t \in [0,10]$ and the domain is $x \in [0,2]$, $y \in [0,1]$.
By examining the FTLE field (which LCS Tool can provide), we position Poincare sections to capture elliptic LCSs.
A script to perform this operation is given in \cref{l:double gyre elliptic LCS} where two Poincare sections are defined (\cref{ll:double gyre poincareSection(1),ll:double gyre poincareSection(2)}).
The free stretching parameter $\lambda$ (\cref{eq:etafields}) is varied over the range $[0.93,1.07]$ with increments of $0.01$ (\cref{ll:double gyre lambdaRange}).
Closed orbits for all predefined $\lambda$ values are computed and the outermost closed orbit is kept as the Lagrangian vortex boundary  (\crefrange{ll:double gyre lambda range}{ll:double gyre elliptic_lcs}).
We have chosen small values for the error tolerances of the integration of the Cauchy-Green strain tensor (\cref{ll:double gyre cgStrainOdeSolverOptions}) and $\lambda$-lines (\cref{ll:double gyre lambdaLineOdeSolverOptions}) to ensure convergence.

@@

In \cref{f:double gyre lambda LCS convergence}, the resolution of the  Cauchy-Green strain tensor is varied from $500 \times 250$ to $1000 \times 500$. The location of the outermost closed $\lambda$-line changes insignificantly, demonstrating convergence. For all tested resolutions, the $\lambda$ values of the outermost closed orbits match.
This also suggests that the lowest tested resolution, $500 \times 250$, is sufficient to identify elliptic LCSs.

\begin{figure}
\centering
\setlength\figurewidth{.4\textwidth}
\import{graphics/double_gyre/}{elliptic_lcs_convergence_500.tikz}
\import{graphics/double_gyre/}{elliptic_lcs_convergence_750.tikz}
\import{graphics/double_gyre/}{elliptic_lcs_convergence_1000.tikz}
\caption{Convergence of closed $\lambda$-lines for increasing main-grid resolution in the double-gyre. The outermost closed $\lambda$-line (bold green line) is the vortex boundary. Top: $500 \times 250$, middle: $750 \times 375$, bottom: $1000 \times 500$.}
\label{f:double gyre lambda LCS convergence}
\end{figure}

@@

\begin{figure}
\centering
\setlength\figurewidth{.35\textwidth}
\import{graphics/double_gyre/}{elliptic_repelling_lcs_details.tikz}
\import{graphics/double_gyre/}{elliptic_attracting_lcs_details.tikz}
\caption{
LCSs in the double gyre.
Resolution is $500 \times 250$.
Elliptic LCSs are green, shrink line LCSs are red and stretch line LCSs are blue.
White dots indicate $\lambda_2$ maxima for shrink lines and $\lambda_1$ minima stretch lines.
FTLE shown in the background.
$\lambda = 1.00$ and $1.04$ for the left and the right gyre.
$\lambda \in [0.93,1.07]$, $\Delta\lambda = 0.01$.
}
\label{f:double gyre lambda hyperbolic LCS}
\end{figure}

@@

The Bickley jet models a meandering zonal jet flanked above and below by
counter rotating vortices. This is an idealized model of geophysical flows such as the Gulf Stream and the polar night jet perturbed by a Rossby wave\citep{castilloNegrete93:_chaot_rossb,beron-vera10:_invar_lagran}.

The velocity is given by $v(x,y,t) = (-\partial_y \psi, \partial_x \psi)$ where
\begin{gather*}
\psi(x,y,t) = \psi_0(x,y) + \psi_1(x,y,t),\\
\psi_0(x,y) = c_3 y - U L_y \tanh\frac{y}{L_y} + \epsilon_3 U L_y \sech^2\frac{y}{L_y} \cos k_3 x,\\
\psi_1(x,y,t) = U L_y \sech^2\frac{y}{L_y} \Re\left[\sum_{n=1}^2 \epsilon_n f_n(t) e^{i k_n x}\right].
\end{gather*}
As a forcing function, we choose a solution running on the chaotic attractor of the damped and forced Duffing oscillator, specifically
\begin{gather*}
\od{\phi_1}{t} = \phi_2,\\
\od{\phi_1}{t} = -0.1 \phi_2 - \phi_1^3 + 11 \cos(t),\\
f_{1,2}(t) = 2.625 \times 10^{-2} \phi_1(t/6.238 \times 10^5)
\end{gather*}
The parameter values we use are: $U = 62.66$, $c_2 = 0.205 U$, $c_3 = 0.461 U$, $L_y = 1.77 \times 10^6$, $\epsilon_1 = 0.0075$, $\epsilon_2 = 0.04$, $\epsilon_3 = 0.3$, $L_x = 6.371 \times 10^6 \pi$, $k_n = 2 n \pi/L_x$, $\sigma_1 = 0.5 k_2 (c_2 - c_3)$, $\sigma_2 = 2 \sigma_1$.

The integration time is $T = 4L_x/U$, an integer multiple of the eddy turnover time.
\Cref{l:Bickley jet elliptic hyperbolic LCS subset} shows the LCS Tool commands in which the chaotically perturbed velocity is defined (\cref{ll:Bickley jet velocity definition}), periodic boundary conditions are imposed in the x-direction (\cref{ll:Bickley jet periodicBc}), and five Poincare sections are defined where we expect coherent vortices (\crefrange{ll:Bickley jet Poincare start}{ll:Bickley jet Poincare end}).
$\lambda$-values for closed orbit detection are varied over the range $[0.80,1.20]$ with a step of $0.01$.

@@

\begin{sloppypar}
The altimeter products used in this work are produced by SSALTO/DUACS and distributed by AVISO, with support from CNES, \url{http://www.aviso.oceanobs.com/duacs}.
\end{sloppypar}

\section*{References}

>>>>

\documentclass{article}

\usepackage{amsmath}
\usepackage[uniquename=init,firstinits,backend=biber,sorting=none]{biblatex}
\usepackage{commath}
\usepackage{fontspec}
\usepackage{gensymb}
\usepackage{import}
\usepackage{listings}
\usepackage[numbered,framed]{mcode}
\usepackage{microtype}
\usepackage{pgfplots}
\usepackage[hidelinks]{hyperref}
\usepackage{cleveref}

@@

\lstloadlanguages{Matlab}
\lstset{
  basicstyle=\small,
  breaklines,
  frame=single,
  float,
  columns=fullflexible,
  numbers=left,
  stepnumber=2,
  backgroundcolor=\color{grey},
  escapeinside={\%@}{\^^M}
}

@@

Lagrangian Coherent Structures (LCSs) are evolving organizing centres of trajectory patterns in non-autonomous dynamical systems\parencite{haller00:_lagran,peacock13:_lagran,haller15:_langr_coher_struc}. Applications of LCSs include oceanic and atmospheric flows\parencite{beron-vera13:_objec_agulh,koh02:_hyper}, biological transport problems\parencite{wilson09:_lagran_reynol,tallapragada11:_lagran,huhn12:_south_indian_ocean_count_madag}, aeronautics\parencite{tang10:_accur_lagran_hong_kong_inter_airpor}, celestial mechanics\parencite{gawlik09:_lagran}, crowd dynamics\parencite{ali07:_lagran_partic_dynam_approac_crowd}, and aperiodically forced mechanical oscillators\parencite{hadjighasem13:_detec_kam}.

\begin{sloppypar}
\textcite{haller01:_distin} proposed that ridges of the finite-time Lyapunov exponent (FTLE) are heuristic indicators of hyperbolic (i.e., repelling and attracting type) LCSs.
A number of examples support this principle\parencite{peacock13:_lagran}.
Equating FTLE ridges with LCSs, however, would create theoretical inconsistencies, as well as false positives and negatives in hyperbolic LCS detection\parencite{haller11:_lagran_coher_struc,norgard12:_secon_lagran_coher_struc}.
In addition, the role of the FTLE field in the accurate detection of elliptic (vortex-type) and parabolic (jet-core type) LCSs has remained an open question (but see \textcite{beron-vera10:_invar_lagran}).
\end{sloppypar}

More recent work has focused on an exact mathematical formulation of the properties defining LCSs\parencite{haller11:_lagran_coher_struc,farazmand12:_comput_lagran,haller12:_geodes_theor_trans_barrier_two_dimen_flows,haller13:_coher_lagran,haller14:_adden_coher_lagran,farazmand13:_attrac_lagran,blazevski14:_hyper_ellip_trans_barrier_three}.
In two-dimensional flows, hyperbolic and parabolic LCSs turn out to be stationary curves of the averaged material shear\parencite{farazmand14:_shearless}, whereas elliptic LCSs are stationary curves of the averaged strain\parencite{haller13:_coher_lagran,haller14:_adden_coher_lagran}.
These variational formulations lead to explicit solutions for LCSs as null geodesics of appropriate Lorentzian metrics.

We present an algorithmic introduction to geodesic LCS detection.
We then review the implementation of this approach in a computational engine called LCS Tool\footnote{LCS Tool is available at: \url{http://github.com/jeixav/LCS-Tool}}.
This engine is a library of MATLAB functions that extract LCSs from two-dimensional unsteady flows. The examples we present form demonstration scripts distributed with LCS Tool.

Other software for LCS detection exists.
ManGen\parencite{lekien03:_time} calculates the FTLE and advects material curves in two-dimensional velocity fields.
It includes a graphical user interface and uses the Message Passing Interface standard for parallel calculations.
Newman\parencite{toit10:_trans} calculates the FTLE using dimension independent code.
It assists ridge extraction of FTLE fields and supports analytic and dataset velocity definitions.
FlowTK\parencite{ameli14:_devel_effic_flexib_pipel_lagran} calculates the FTLE in two and three-dimensions on Cartesian and unstructured grids.
The NVIDIA CUDA parallel computing platform is used for fast computations and Kitware's ParaView data analysis and visualization application is used for a user interface.

@@

Next we consider positions of material lines at time $t_0$ that prevail as most repelling or attracting material lines (or \emph{hyperbolic LCSs}) over a time interval $[t_0,t] \subset [t_-,t_+]$.
\textcite{farazmand14:_shearless} argue that hyperbolic LCSs are stationary curves of the averaged shear functional
\begin{gather*}
Q(\gamma) = \frac{1}{\sigma} \int_0^\sigma \frac{\langle r'(s),D_{t_0}^t(r(s)) r'(s)\rangle}{\sqrt{\langle r'(s),C_{t_0}^t(r(s)) r'(s)\rangle\langle r'(s),r'(s)\rangle}}\dif s,\\
D_{t_0}^t = \frac12 [C_{t_0}^t \Omega - \Omega C_{t_0}^t],
\end{gather*}
obtained by averaging the Lagrangian shear arising over $[t_0,t]$ along material lines parametrized as $r(s)$ with $s \in [0,\sigma]$.
More precisely, hyperbolic LCSs are stationary curves of $Q(\gamma)$ with respect to fixed-endpoint perturbations.
We note that parabolic LCSs (Lagrangian jet cores) are also stationary curves of $Q(\gamma)$, but under variable endpoint perturbations (cf. \textcite{farazmand14:_shearless}).

@@

\begin{sloppypar}
Shrink lines and stretch lines coincide with the null geodesics of the Lorentzian metric $h(u,v) = \left\langle u,D_{t_0}^t(r) v \right\rangle$.
For this reason, we refer to the computation of hyperbolic LCSs as strongest normally-repelling or normally-attracting curves of \cref{eq:etafields} as \emph{geodesic detection of hyperbolic LCSs}.
\end{sloppypar}

@@

\begin{table}
\begin{enumerate}
\item Define a velocity field, \lstinline!dx = derivative(t,x,p)!
\item Compute Cauchy-Green strain tensor invariants, \lstinline![v,d] = eig_cgStrain(derivative)!
\item Define a range of $\lambda$ values, \lstinline!lambda!
\item Define Poincare sections, \lstinline!ps!
\item Detect elliptic LCSs, \lstinline!closedLambdaLine = poincare_closed_orbit_range(v,d,lambda,ps)! and \lstinline!ellipticLcs = elliptic_lcs(closedLambdaLine)!
\item Define a local maximization/minimization distance for hyperbolic LCSs, \lstinline!lmd!
% FIXME seed_curves_from_lambda_max is a poor name for a function that detects hyperbolic LCSs
\item Detect hyperbolic LCSs, \lstinline!hyperbolicLine = seed_curves_from_lambda_max(v,d,ellipticLcs,lmd)! and \lstinline!hyperbolicLcs = remove_strain_in_elliptic(hyperbolicLine,ellipticLcs)!
\end{enumerate}
\caption{Overview of sequence of computations to detect LCSs with LCS Tool functions.}
\label{t:LCS Tool overview}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Resolution & Runtime (s) & Memory (MB)\tabularnewline
\hline \hline
$500 \times 251$ & 29 & 55\tabularnewline
\hline
$750 \times 376$ & 63 & 123\tabularnewline
\hline
$1000 \times 501$ & 112 & 223\tabularnewline
\hline
\end{tabular}
\end{center}
\caption{
Runtime and memory use associated with vector integration at different resolutions.
Results are from the function \lstinline!eig_cgStrain! applied to the double gyre presented in \cref{sec:double gyre}.
Results obtained with an Intel Core i3-4130 processor.
}
\label{t:vector integration memory}
\end{table}

The Cauchy-Green strain tensor can have point singularities, i.e., points where $C_{t_0}^t(x_0)$ has repeated eigenvalues.
At these points the eigenvectors $\xi_1(x_0)$ and $\xi_2(x_0)$ are no longer well-defined.
This generically arises at a finite set of isolated points within the computational domain\parencite{delmarcelle94}, and hence lie off the computational grid with probability one.

@@

% FIXME lambda_line and poincare_closed_orbit have been removed from demo scripts. Now use poincare_closed_orbit_range and elliptic_lcs.
\begin{table*}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline
\multicolumn{2}{|p{.95\textwidth}|}
{\lstinline![shearline.etaPos,shearline.etaNeg] = lambda_line(cgEigenvector,cgEigenvalue,lambda)!}\tabularnewline
\hline \hline
\lstinline!cgEigenvector! & array of Cauchy-Green strain eigenvectors\tabularnewline
\hline
\lstinline!cgEigenvalue! & array of Cauchy-Green strain eigenvalues\tabularnewline
\hline
\lstinline!lambda! & scalar lambda value in \cref{eq:eta}\tabularnewline
\hline \hline
\multicolumn{2}{|p{.95\textwidth}|}{\lstinline![closedOrbits,orbits] = poincare_closed_orbit_multi(domain,resolution,shearline,PSList)!}\tabularnewline
\hline \hline
\lstinline!domain! & array to define flow domain\tabularnewline
\hline
\lstinline!resolution! & $1 \times 2$ array to define main grid resolution for Cauchy-Green strain tensor\tabularnewline
\hline
\lstinline!shearline! & structure of arrays of $\eta_+$ and $\eta_-$ values on main grid\tabularnewline
\hline
\lstinline!PSList! & user-defined structure for Poincare section end-points, number of $\lambda$-lines launched from Poincare section, and maximum closed $\lambda$-line length\tabularnewline
\hline
\lstinline!nBisection! & optional number of bisection steps to refine zero crossings of Poincare map. Default: 5.\tabularnewline
\hline
\lstinline!dThresh! & optional threshold to discard discontinuous zero crossings of Poincare map. Default: $10^{-2}$.\tabularnewline
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify $\lambda$-line integration parameters\tabularnewline
\hline
\lstinline!periodicBc! & optional $1 \times 2$ logical array to specify periodic boundary conditions. Default: \lstinline![false,false]!.\tabularnewline
\hline
\end{tabular}
\caption{
Syntax for LCS Tool elliptic LCS functions.
Both functions are called by the function \lstinline!poincare_closed_orbit_range!.
}
\label{t:Elliptic LCS functions}
\end{table*}

The first step is to define the position of Poincare sections in regions where closed $\lambda$-lines are expected based on a visual analysis of the orbit structure of the $\eta_\pm^\lambda$ vector field.
The Poincare section is to be oriented such that the first endpoint is close to the centre of the expected Lagrangian vortex, and the second endpoint is outside this vortex.
% FIXME Move citation to conclusions, as a candidate for future work.
There is no automated procedure implemented in LCS Tool for the positioning of Poincare sections (see however \textcite{karrasch14:_autom_lagran} for a recently developed algorithm).
Additionally, the number of lambda-lines launched from the Poincare section, \lstinline!poincareSection.numPoints!, must be defined.
A reasonable default value is 100.

@@

\begin{table}
\begin{center}
\begin{tabular}{|c|c|}
\hline
Demo & Runtime (min)\tabularnewline
\hline \hline
double\_gyre/elliptic\_hyperbolic\_lcs & 50\tabularnewline
\hline
bickley\_jet/elliptic\_hyperbolic\_lcs\_details & 975\tabularnewline
\hline
ocean\_dataset/elliptic\_hyperbolic\_lcs\_details & 127\tabularnewline
\hline
\end{tabular}
\end{center}
\caption{
Runtime of some demo scripts.
Results obtained with an Intel Core i3-4130 processor with eight gigabytes of memory.
}
\label{t:demo runtimes}
\end{table}

@@

The double gyre is a model for a time-dependent two gyre system observed in geophysical flows \parencite{shadden05:_defin_lagran_lyapun}. The model consists of two counter rotating sinusoidal vortices with a harmonically oscillating line in-between. Lagrangian particle motions satisfy the non-autonomous dynamical system
\begin{equation}
\begin{split}
\od{x}{t} = -\pi A \sin[\pi f(x,t)] \cos(\pi y),\\
\od{y}{t} = \pi A \cos[\pi f(x,t)] \sin(\pi y) \pd{f(x,t)}{x},\\
f(x,t) = \epsilon \sin(\omega t) x^2 + [1 - 2 \epsilon \sin(\omega t)] x.
\end{split}
\label{eq:double gyre derivative equations}
\end{equation}
The MATLAB function describing this velocity field is given in \cref{l:double gyre derivative}, specifying the right hand side of the particle ordinary differential equation in a way that supports vectorized integration.

@@

In what follows, the parameter values are: $A = 0.1$, $\epsilon = 0.1$, $\omega = \pi/5$.
The flow timespan is $t \in [0,10]$ and the domain is $x \in [0,2]$, $y \in [0,1]$.
By examining the FTLE field (which LCS Tool can provide), we position Poincare sections to capture elliptic LCSs.
A script to perform this operation is given in \cref{l:double gyre elliptic LCS} where two Poincare sections are defined (\cref{ll:double gyre poincareSection(1),ll:double gyre poincareSection(2)}).
The free stretching parameter $\lambda$ (\cref{eq:etafields}) is varied over the range $[0.93,1.07]$ with increments of $0.01$ (\cref{ll:double gyre lambdaRange}).
Closed orbits for all $\lambda$ values are computed and the outermost closed orbit is kept as the Lagrangian vortex boundary  (\crefrange{ll:double gyre lambda range}{ll:double gyre elliptic_lcs}).
We have chosen small values for the error tolerances of the integration of the Cauchy-Green strain tensor (\cref{ll:double gyre cgStrainOdeSolverOptions}) and $\lambda$-lines (\cref{ll:double gyre lambdaLineOdeSolverOptions}) to ensure convergence.

@@

In \cref{f:double gyre lambda LCS convergence}, the resolution of the  Cauchy-Green strain tensor is varied from $500 \times 250$ to $1000 \times 500$. The location of the outermost closed $\lambda$-line changes insignificantly, demonstrating convergence. For all tested resolutions, the $\lambda$ values of the outermost closed orbits match.
% FIXME Should reduce lowest resolution used to demonstrate problems if selected resolution is too low
This also suggests that the lowest tested resolution, $500 \times 250$, is sufficient to identify elliptic LCSs.

\begin{figure}
\centering
\setlength\figurewidth{.875\textwidth}
\import{graphics/double_gyre/}{elliptic_lcs_convergence_500.tikz}
\import{graphics/double_gyre/}{elliptic_lcs_convergence_750.tikz}
\import{graphics/double_gyre/}{elliptic_lcs_convergence_1000.tikz}
\caption{Convergence of closed $\lambda$-lines for increasing main-grid resolution in the double-gyre. The outermost closed $\lambda$-line (bold green line) is the vortex boundary. Top: $500 \times 250$, middle: $750 \times 375$, bottom: $1000 \times 500$.}
\label{f:double gyre lambda LCS convergence}
\end{figure}

@@

\begin{figure}
\centering
\setlength\figurewidth{.8\textwidth}
\import{graphics/double_gyre/}{elliptic_repelling_lcs_details.tikz}
\import{graphics/double_gyre/}{elliptic_attracting_lcs_details.tikz}
\caption{
LCSs in the double gyre.
Resolution is $500 \times 250$.
Elliptic LCSs are green, shrink line LCSs are red and stretch line LCSs are blue.
White dots indicate $\lambda_2$ maxima for shrink lines and $\lambda_1$ minima stretch lines.
FTLE shown in the background.
$\lambda = 1.00$ and $1.04$ for the left and the right gyre.
$\lambda \in [0.93,1.07]$, $\Delta\lambda = 0.01$.
}
\label{f:double gyre lambda hyperbolic LCS}
\end{figure}

@@

The Bickley jet models a meandering zonal jet flanked above and below by
counter rotating vortices. This is an idealized model of geophysical flows such as the Gulf Stream and the polar night jet perturbed by a Rossby wave\parencite{castilloNegrete93:_chaot_rossb,beron-vera10:_invar_lagran}.

The velocity is given by $v(x,y,t) = (-\partial_y \psi, \partial_x \psi)$ where
\begin{gather*}
\psi(x,y,t) = \psi_0(x,y) + \psi_1(x,y,t),\\
\psi_0(x,y) = c_3 y - U L_y \tanh\frac{y}{L_y} + \epsilon_3 U L_y \sech^2\frac{y}{L_y} \cos k_3 x,\\
\psi_1(x,y,t) = U L_y \sech^2\frac{y}{L_y} \Re\left[\sum_{n=1}^2 \epsilon_n f_n(t) e^{i k_n x}\right].
\end{gather*}
As a forcing function, we choose a solution running on the chaotic attractor of the damped and forced Duffing oscillator, specifically
\begin{gather*}
\od{\phi_1}{t} = \phi_2,\\
\od{\phi_2}{t} = -0.1 \phi_2 - \phi_1^3 + 11 \cos(t),\\
f_{1,2}(t) = 2.625 \times 10^{-2} \phi_1(t/6.238 \times 10^5)
\end{gather*}
The parameter values we use are: $U = 62.66$, $c_2 = 0.205 U$, $c_3 = 0.461 U$, $L_y = 1.77 \times 10^6$, $\epsilon_1 = 0.0075$, $\epsilon_2 = 0.04$, $\epsilon_3 = 0.3$, $L_x = 6.371 \times 10^6 \pi$, $k_n = 2 n \pi/L_x$, $\sigma_1 = 0.5 k_2 (c_2 - c_3)$, $\sigma_2 = 2 \sigma_1$.

The integration time is $T = 4L_x/U$, an integer multiple of the eddy turnover time.
\Cref{l:Bickley jet elliptic hyperbolic LCS subset} shows the LCS Tool commands in which the chaotically perturbed velocity is defined (\crefrange{ll:Bickley jet velocity definition start}{ll:Bickley jet velocity definition end}), periodic boundary conditions are imposed in the x-direction (\cref{ll:Bickley jet periodicBc}), and five Poincare sections are defined where we expect coherent vortices (\crefrange{ll:Bickley jet Poincare start}{ll:Bickley jet Poincare end}).
$\lambda$-values for closed orbit detection are varied over the range $[0.80,1.20]$ with a step of $0.01$.

@@

The altimeter products used in this work are produced by SSALTO/DUACS and distributed by AVISO, with support from CNES, \url{http://www.aviso.oceanobs.com/duacs}.

% FIXME Citations without pages (karrasch14:_autom_lagran) do not list publication year
\printbibliography
